\Large\bf FlexibleSUSY --- A spectrum generator generator for supersymmetric models

### 2013-11-03 doc/paper.tex
more references and copied over details on two-scale solver.

  \begin{abstract}
   Here we introduce \fs, a MATHEMATICA package which generates a C++
   spectrum generator for any SUSY model, created with modularity and
   speed in mind.  \fs makes use of the existing \sarah package to
   obtain self energies, tadpoles corrections and renormalisation
   group equations in any model.  \fs then generates C++ code for
   calculating spectra with a structure that is designed to be
   flexible. The user specifies the model supplying the superpotential,
   gauge structure and particle content in a \sarah model file. The
   user should also specify the boundary conditions in a separate \fs
   model file. {\color{red}( Do we want to keep this separate and in
     different places, add a script to put a SARAH model file in the
     right place or combine them?)}  Flexible SUSY then translates
   these into C++ code along with numerical routines for solving the
   renormalisation group equations, one loop electroweak symmetry
   breaking conditions and a solver routine for simultaneously solving
   for the spectrum.  The modular structure of the generated code
   allows for any individual component to be replaced with an
   alternative if available.  Indeed \fs has been intentionally
   designed to grow as alternative solvers and calculators are added.
  \end{abstract}
\begin{keyword}
sparticle, 
supersymmetry, 
Higgs
\PACS 12.60.Jv
\PACS 14.80.Ly
\end{keyword}
\end{frontmatter}

@@

\subsection{The Program}
\fs is a MATHEMATICA package designed to create a fast and easily
adaptable spectrum generator in C++ for any SUSY model and some
non-supersymmetric models {\color{red} can we?}.  \fs makes use of the
existing \sarah package to obtain one-loop self energies, one-loop
tadpoles corrections and two-loop renormalisation group
equations(RGEs) in any model.  \fs then converts these algerbraic
expressions from \sarah into C++ code using the C++11 standard. \fs
uses some parts of SOFTSUSY\cite{Allanach:2001kg}, the very fast Eigen
library \cite{eigen}, as well as the gnu scientific library and the
boost library to create numerical routines which then use these
routines to solve the RGEs and EWSB conditions which will be used in
the spectrum generator. The main solver routine of \fs then calls
these routines to find simultaneous solutions for the EWSB, low energy
data, and other, user supplied, boundary conditions (e.g. the usual
mSUGRA universality conditions) and calculates the mass spectrum.

@@

\section{Stucture of the spectrum generator}
\subsection{boundary conditions}
Different types, how we match to low energy data.  reference EWSB or merge that section here.
mention two loop qcd parts from softsusy.
\subsection{RGEs}
The two loop renormalisation group equations are generated using command ?
Even in the simplest supersymmetric model the RGEs form a very large set of coupled non-linear differential equations.  To solve these and thereby evolve the parameters between two different scales \fs has... 
\subsection{tree level spectrum}
The diagonalised tree level $\overline{DR}$ masses are obtained from expressions...
The numerical diagonalisation is performed by... with an accuracy set by...   
\subsection{Electroweak symmetry breaking}
At tree level the EWSB conditions are written...
The one loop electroweak symmetry breaking are solved using the gsl routines finding multiple roots, \code{gsl_multiroot_fsolver}.  Iteration. For example in the MSSM \fs generates.. show code.
 If higher accuracy is required additional routines with higher order corrections can be added in a very simple way.  For example in the MSSM by default \fs adds two loop Higgs FORTRAN routines supplied by P.~Slavich from \cite{slavich-degrassi} to add two loop corrections of $\oatas$, $\oabas$, $\oatq$, $\oabatau$, 
$\oabq$, $\oatauq$ and $\oatab$  as follows...  
Something similar is done for the NMSSM, but in this case the NMSSM $\oatas$, $\oabas$ pieces come from \cite{Degrassi:2009yq}.
 Corrections from other sources can be added to the generated code by...
\subsection{two scale fixed point iteration}
\subsection{pole masses}
With the solver routine finished and convergence achieved, the $\overline{DR}$ parameters consistent with one loop EWSB, low energy data and all user supplied boundary conditions are then known at any scale between...

@@

>>>>

  \begin{abstract}
   Here we introduce \fs, a MATHEMATICA package which generates a C++
   spectrum generator for any SUSY model, created with modularity and
   speed in mind.  \fs makes use of the existing \sarah package
   to obtain self energies, tadpoles corrections and renormalisation
   group equations in any model.  \fs then generates C++ code for
   calculating spectra with a structure that is designed to be
   flexible. The user specifies the model supplying the
   superpotential, gauge structure and particle content in a \sarah
   model file. The user should also specify the boundary conditions in
   a separate \fs model file. {\color{red}( Do we want to keep this
     separate and in different places, add a script to put a \sarah
     model file in the right place or combine them?)}  Flexible SUSY
   then translates these into C++ code along with numerical routines
   for solving the renormalisation group equations, one loop
   electroweak symmetry breaking conditions and a solver routine for
   simultaneously solving for the spectrum.  The modular structure of
   the generated code allows for any individual component to be
   replaced with an alternative if available.  Indeed \fs has been
   intentionally designed to grow as alternative solvers and
   calculators are added.
  \end{abstract}
\begin{keyword}
sparticle, 
supersymmetry, 
Higgs
\PACS 12.60.Jv
\PACS 14.80.Ly
\end{keyword}
\end{frontmatter}

@@

\subsection{The Program}
\fs is a MATHEMATICA package designed to create a fast and easily
adaptable spectrum generator in C++ for any SUSY model and some
non-supersymmetric models {\color{red} can we?}.  \fs makes use of the
existing \sarah package \cite{Staub:2010ty,Staub:2009bi,Staub:2010jh,Staub:2012pb,Staub:2013tta} to obtain one-loop self energies, one-loop
tadpoles corrections and two-loop renormalisation group
equations(RGEs) in any model.  \fs then converts these algerbraic
expressions from \sarah into C++ code using the C++11 standard. \fs
uses some parts of SOFTSUSY\cite{Allanach:2001kg}, the very fast Eigen
library \cite{eigen}, as well as the gnu scientific library and the
boost library to create numerical routines which then use these
routines to solve the RGEs and EWSB conditions which will be used in
the spectrum generator. The main solver routine of \fs then calls
these routines to find simultaneous solutions for the EWSB, low energy
data, and other, user supplied, boundary conditions (e.g. the usual
mSUGRA universality conditions) and calculates the mass spectrum.

@@

\section{Stucture of the spectrum generator}
\subsection{boundary conditions}
Different types, how we match to low energy data.  reference EWSB or merge that section here.
mention two loop qcd parts from softsusy.
\subsection{RGEs}
The two loop renormalisation group equations are generated using command ?
Even in the simplest supersymmetric model the RGEs form a very large set of coupled non-linear differential equations.  To solve these and thereby evolve the parameters between two different scales \fs has... 
\subsection{tree level spectrum}
The diagonalised tree level $\overline{DR}$ masses are obtained from expressions...
The numerical diagonalisation is performed by... with an accuracy set by...   
\subsection{Electroweak symmetry breaking}
At tree level the EWSB conditions are written...
The one loop electroweak symmetry breaking are solved using the gsl routines finding multiple roots, \code{gsl_multiroot_fsolver}.  Iteration. For example in the MSSM \fs generates.. show code.
 If higher accuracy is required additional routines with higher order corrections can be added in a very simple way.  For example in the MSSM by default \fs adds two loop Higgs FORTRAN routines supplied by P.~Slavich from \cite{slavich-degrassi} to add two loop corrections of $\oatas$, $\oabas$, $\oatq$, $\oabatau$, 
$\oabq$, $\oatauq$ and $\oatab$  as follows...  
Something similar is done for the NMSSM, but in this case the NMSSM $\oatas$, $\oabas$ pieces come from \cite{Degrassi:2009yq}.
 Corrections from other sources can be added to the generated code by...
\subsection{two scale fixed point iteration}
The default solver which calls the above routines above and finds the full set of $\overline{DR}$ masses in the model consistent with all constraints is the two-scale solver.  This can be selected by... but if no selection is made this will be used anyway as default.

This approach is decribed in \cite{Barger:1993gh} for the MSSM and is widely implemented in SUSY spectrum generators.

The stucture of the two-scale algorithm used in \fs is as follows,
%
Initial guess:
\begin{enumerate}
\item Guess gauge couplings $g_{1,2,3}$ at $m_Z$
\item Apply user-defined low-scale constraint (\code{LowScaleInput})
\item Guess Yukawa couplings $y^{u,d,e}_{ij}$ at $m_Z$
\item Run model to the high-scale (\code{HighScaleFirstGuess})
\item Apply high-scale constraint (\code{HighScaleInput})
\item Run model to the low-scale (\code{LowScaleFirstGuess})
\item Solve EWSB eqs.\ at the tree-level
\item Calculate \DRbar\ masses
\end{enumerate}
%
Solving the renormalization group equations:
\begin{enumerate}
\item \label{rge-step-one} Run model to the low-scale (\code{LowScale})
  \begin{enumerate}
  \item Calculate \DRbar\ masses
  \item Recalculate low-scale
  \item Calculate \DRbar\ gauge and Yukawa couplings $g_{1,2,3}$, $y^{u,d,e}_{ij}$
  \item Apply user-defined low-scale constraint (\code{LowScaleInput})
  \end{enumerate}
\item Run model to the high-scale (\code{HighScale})
  \begin{enumerate}
  \item Recalculate high-scale
  \item Apply high-scale constraint (\code{HighScaleInput})
  \end{enumerate}
\item Run model to the susy-scale (\code{SUSYScale})
  \begin{enumerate}
  \item Calculate \DRbar\ masses
  \item Recalculate low-scale
  \item Apply susy-scale constraint (\code{SUSYScaleInput})
  \item Solve EWSB eqs.\ at the one-loop level
  \end{enumerate}
\item If not converged yet, goto \ref{rge-step-one}
\end{enumerate}
%
Calculating the particle spectrum:
\begin{enumerate}
\item Run model to the susy-scale
\item Calculate the pole masses
\item Run model to the parameter output scale (SLHA block MODSEL 12)
\end{enumerate}

\begin{figure}[tb]
  \centering
  \tikzumlset{fill class=white}
  \begin{tikzpicture}
    \umlclass[x=0, y=0, type=abstract]{Beta\_function}{
      -- scale : double\\
      -- loops : unsigned\\
      -- numPars : unsigned
    }{
      + \umlvirt{display() : const Eigen::ArrayXd} \\
      + \umlvirt{set(v : const Eigen::ArrayXd\&) : void}\\
      + \umlvirt{beta() : Eigen::ArrayXd}\\
      + run\_to(scale : double, eps : double) : void\\
    }
    \umlclass[x=8, y=-12, template={T}]{MSSM}{}{}
    \umlclass[x=8, y=-8, type=abstract]{Two\_scale\_model}{}{
      + \umlvirt{calculate\_spectrum() : void}\\
      + \umlvirt{run\_to(scale : double, eps : double) : void}\\
      + \umlvirt{set\_precision(precision : double) : void}
    }
    \umlclass[x=0, y=-4]{MSSM\_susy\_parameters}{}{
      + display() : const Eigen::ArrayXd \\
      + set(v : const Eigen::ArrayXd\&) : void\\
      + beta() : Eigen::ArrayXd
    }
    \umlclass[x=0, y=-8]{MSSM\_soft\_parameters}{}{
      + display() : const Eigen::ArrayXd \\
      + set(v : const Eigen::ArrayXd\&) : void\\
      + beta() : Eigen::ArrayXd
    }
    \umlclass[x=0, y=-12]{MSSM$<$Two\_scale$>$}{}{
      + calculate\_spectrum() : void\\
      + run\_to(scale : double, eps : double) : void\\
      + set\_precision(precision : double) : void
    }
   \umlinherit{MSSM\_susy\_parameters}{Beta\_function}
   \umlinherit{MSSM\_soft\_parameters}{MSSM\_susy\_parameters}
   \umlinherit{MSSM$<$Two\_scale$>$}{MSSM\_soft\_parameters}
   \umlinherit{MSSM$<$Two\_scale$>$}{Two\_scale\_model}
   \umldep[arg1={$<<$bind$>>$}, mult1={T $\rightarrow$ Two\_scale}, pos1=0.5]{MSSM$<$Two\_scale$>$}{MSSM}
  \end{tikzpicture}
  \caption{Two-scale model class hierarchy}
  \label{fig:two-scale-model-class-hierarchy}
\end{figure}

\begin{figure}[tb]
  \centering
  \tikzumlset{fill class=white}
  \begin{tikzpicture}
    \umlclass[x=0, y=0, type=abstract]{Two\_scale\_model}{}{
      + \umlvirt{calculate\_spectrum() : void}\\
      + \umlvirt{run\_to(scale : double, eps : double) : void}\\
      + \umlvirt{set\_precision(precision : double) : void}
    }
    \umlclass[x=0, y=-4]{RGFlow$<$Two\_scale$>$}{}{
      + solve() : void
    }
    \umlclass[x=8, y=0]{Constraint$<$Two\_scale$>$}{}{
      + \umlvirt{apply() : void}\\
      + \umlvirt{get\_scale() : double}\\
      + \umlvirt{set\_model(model : Two\_scale\_model*) : void}
    }
    \umlclass[x=8, y=4, template={T}]{Constraint}{}{}
    \umlclass[x=0, y=-7]{Initial\_guesser$<$Two\_scale$>$}{}{
      + \umlvirt{guess() : void}
    }
    \umlclass[x=0, y=-11, template={T}]{Initial\_guesser}{}{}
    \umlclass[x=8, y=-7]{Convergence\_tester$<$Two\_scale$>$}{}{
      + \umlvirt{accuracy\_goal\_reached() : bool}\\
      + \umlvirt{max\_iterations() : unsigned}
    }
    \umlclass[x=8, y=-11, template={T}]{Convergence\_tester}{}{}
    \umlclass[x=8, y=-4, template={T}]{RGFlow}{}{}
    \umldep[arg1={$<<$bind$>>$}, mult1={T $\rightarrow$ Two\_scale}, pos1=0.5]{RGFlow$<$Two\_scale$>$}{RGFlow}
    \umldep[mult1=1..*]{RGFlow$<$Two\_scale$>$}{Two\_scale\_model}
    \umldep[mult1=1..*]{RGFlow$<$Two\_scale$>$}{Constraint$<$Two\_scale$>$}
    \umldep{RGFlow$<$Two\_scale$>$}{Initial\_guesser$<$Two\_scale$>$}
    \umldep{RGFlow$<$Two\_scale$>$}{Convergence\_tester$<$Two\_scale$>$}
    \umldep[mult1={$<<$bind$>>$}, arg1={T $\rightarrow$ Two\_scale}, pos1=0.5]{Initial\_guesser$<$Two\_scale$>$}{Initial\_guesser}
    \umldep[mult1={$<<$bind$>>$}, arg1={T $\rightarrow$ Two\_scale}, pos1=0.5]{Convergence\_tester$<$Two\_scale$>$}{Convergence\_tester}
    \umldep[arg1={$<<$bind$>>$}, mult1={T $\rightarrow$ Two\_scale}, pos1=0.5]{Constraint$<$Two\_scale$>$}{Constraint}
  \end{tikzpicture}
  \caption{Two-scale renormalization group solver class hierarchy}
  \label{fig:two-scale-rgflow-class-hierarchy}
\end{figure}
{\color{red} copied and pasted from latest manual -- not sure how best to maintain this, could split maual up into tiny bits, but we may also want some differences?}
\subsection{pole masses}
With the solver routine finished and convergence achieved, the $\overline{DR}$ parameters consistent with one loop EWSB, low energy data and all user supplied boundary conditions are then known at any scale between...

@@

\bibitem{Barger:1993gh} 
  V.~D.~Barger, M.~S.~Berger and P.~Ohmann,
  %``The Supersymmetric particle spectrum,''
  Phys.\ Rev.\ D {\bf 49}, 4908 (1994)
  [hep-ph/9311269].
  %%CITATION = HEP-PH/9311269;%%
  %365 citations counted in INSPIRE as of 03 Nov 2013
